Verifying Cache Coherence Protocols for an Infinite Family of Cache Hierarchies

### 2014-02-06 paper/Intuition.tex

Given these invariants, we will show how Theorem~\fullref{storeAtomicity}
follows.
%From the transitions of Figure \ref{trans}, a store may happen only
%when a leaf cache is in \Mo{} state for that
%address (Invariant~\textit{processReq}).
Leaf caches cannot be ancestors of any other node, so by Invariant~
\fullref{nonAncestorCompatible}, if a leaf cache is in \Mo{} state, all other
leaf caches will be in \In{} state, blocking simultaneous
load and store requests (because of Invariant~\ref{processReq}) for the same
address at the same time at different leaf caches. Thus, we have proved
Definition~\fullref{uniqStTimes}.

It is easy to see that Definition~\fullref{uniqRespLabels}, Definition~
\fullref{localOrdering}, and Definition~\fullref{allPrevious} directly follow
from Invariant~\fullref{noReqAgain}.  As leaf caches have no children, one in
at least \Sh{} state for address $a$ is \clean{} for that address. Since
processing a load request requires the cache to be at least in the \Sh{} state,
its response, which supplies the data for that address present in the cache,
supplies the latest value by Invariant~\fullref{latestValue}. Thus, all the
conditions for Theorem~\fullref{storeAtomicity} are satisfied given Invariants~
\ref{latestValue} to \ref{noReqAgain}.

@@

While Invariants~\fullref{processReq} and \fullref{noReqAgain} can be seen
directly from the MSI transitions, correctness of Invariants~
\fullref{nonAncestorCompatible} and \fullref{latestValue} requires further
decomposition. We present 4 more invariants that are helpful in the proof.

@@

We prove Invariant~\fullref{latestValue} by strong induction on time. Consider
two cases: (a) cache $c$ is already \clean{} for an address $a$ at time $t$ or
(b) the contrary.  If \clean{}, then it has the latest value at time $t$. Since
none of its children's directory entries can be in \Mo{}, all its descendants
should be no greater than \Sh{} for that address (because of Invariant~
\fullref{conservative} and the fact that the $\ge$ relation is preserved for the
state of a cache and that of its ancestor). Any non-ancestor of $c$ and
non-descendant of $c$ must be compatible (as discussed in the previous
paragraph).  Since $c$ is \clean, it is at least in the \Sh{} state, so
compatibility means that non-descendants cannot be in \Mo{} state. All of
$c$'s ancestors are non-leaves, so they cannot process any store requests.
Therefore, no cache save $c$ may have processed a store request, which means that
$c$ contains the latest value for address $a$ at time $t+1$ if it was \clean{}
at time $t$.

If $c$ was not clean at time $t$, it must have received data from either its
parent or its child, and the sender is \clean{} by Invariant~
\ref{noTransitWrite}. By the induction hypothesis, it means that the sender has
the latest value for the address. Again, because of
Invariant~\ref{noTransitWrite}, no store request can be processed when the data
is in transit, which means $c$ will have the latest value for address $a$ at
time $t+1$, establishing Invariant~\fullref{latestValue}.

@@

Figure~\ref{resps} shows an example of the effect of Invariants~
\fullref{stateChange}, \fullref{noCross} and \fullref{respFIFO}. The directory
entry of the parent or the state of the child changes on sending and receiving
a response and the new state is shown at the points of sending or receiving
responses.
Both the cache and its parent follow the same sequence of state and directory
entry changes, but the receiver always lags slightly behind the sender.
The receiver cannot send any response before receiving because of
\fullref{noCross}, thereby eliminating any scenarios where the lag complicates
reasoning.

@@

Proving the other 3 invariants in Section~\ref{secondLevel} requires using
more properties of the MSI transistions of Figure~\ref{trans}, though the argument is
straightforward. Analyzing a parent-child pair will suffice in proving Invariants~
\fullref{localCompatibility1} and \fullref{localCompatibility2}. Proof of Invariant~
\fullref{noTransitWrite} is similar to the proof of Invariant~
\fullref{nonAncestorCompatible}.

>>>>

Given these invariants, we will show how Theorem~\fullref{storeAtomicity}
follows.
%From the transitions of Figure \ref{trans}, a store may happen only
%when a leaf cache is in \Mo{} state for that
%address (Invariant~\textit{processReq}).
Leaf caches cannot be ancestors of any other node, so by Invariant~\fullref{nonAncestorCompatible}, if a leaf cache is in \Mo{} state, all other
leaf caches will be in \In{} state, blocking simultaneous
load and store requests (because of Invariant~\ref{processReq}) for the same
address at the same time at different leaf caches. Thus, we have proved
Definition~\fullref{uniqStTimes}.

It is easy to see that Definition~\fullref{uniqRespLabels}, Definition~
\fullref{localOrdering}, and Definition~\fullref{allPrevious} directly follow
from Invariant~\fullref{noReqAgain}.  As leaf caches have no children, one in
at least \Sh{} state for address $a$ is \clean{} for that address. Since
processing a load request requires the cache to be at least in the \Sh{} state,
its response, which supplies the data for that address present in the cache,
supplies the latest value by Invariant~\fullref{latestValue}. Thus, all the
conditions for Theorem~\fullref{storeAtomicity} are satisfied given Invariants~\ref{latestValue} to \ref{noReqAgain}.

@@

While Invariants~\fullref{processReq} and \fullref{noReqAgain} can be seen
directly from the MSI transitions, correctness of Invariants~\fullref{nonAncestorCompatible} and \fullref{latestValue} requires further
decomposition. We present 4 more invariants that are helpful in the proof.

@@

We prove Invariant~\fullref{latestValue} by strong induction on time. Consider
two cases: (a) cache $c$ is already \clean{} for an address $a$ at time $t$ or
(b) the contrary.  If \clean{}, then it has the latest value at time $t$. Since
none of its children's directory entries can be in \Mo{}, all its descendants
should be no greater than \Sh{} for that address (because of Invariant~\fullref{conservative} and the fact that the $\ge$ relation is preserved for the
state of a cache and that of its ancestor). Any non-ancestor of $c$ and
non-descendant of $c$ must be compatible (as discussed in the previous
paragraph).  Since $c$ is \clean, it is at least in the \Sh{} state, so
compatibility means that non-descendants cannot be in \Mo{} state. All of
$c$'s ancestors are non-leaves, so they cannot process any store requests.
Therefore, no cache save $c$ may have processed a store request, which means that
$c$ contains the latest value for address $a$ at time $t+1$ if it was \clean{}
at time $t$.

If $c$ was not clean at time $t$, it must have received data from either its
parent or its child, and the sender is \clean{} by Invariant~\ref{noTransitWrite}. By the induction hypothesis, it means that the sender has
the latest value for the address. Again, because of
Invariant~\ref{noTransitWrite}, no store request can be processed when the data
is in transit, which means $c$ will have the latest value for address $a$ at
time $t+1$, establishing Invariant~\fullref{latestValue}.

@@

Figure~\ref{resps} shows an example of the effect of Invariants~\fullref{stateChange}, \fullref{noCross} and \fullref{respFIFO}. The directory
entry of the parent or the state of the child changes on sending and receiving
a response and the new state is shown at the points of sending or receiving
responses.
Both the cache and its parent follow the same sequence of state and directory
entry changes, but the receiver always lags slightly behind the sender.
The receiver cannot send any response before receiving because of
\fullref{noCross}, thereby eliminating any scenarios where the lag complicates
reasoning.

@@

Proving the other 3 invariants in Section~\ref{secondLevel} requires using
more properties of the MSI transistions of Figure~\ref{trans}, though the argument is
straightforward. Analyzing a parent-child pair will suffice in proving Invariants~\fullref{localCompatibility1} and \fullref{localCompatibility2}. Proof of Invariant~\fullref{noTransitWrite} is similar to the proof of Invariant~\fullref{nonAncestorCompatible}.
